Draft final report\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-10-28 draftfinalreport.tex
New benefit cost and final report draft
Using Mick's values and new simulations, added new benefit-cost
analyses, incorporating the management costs.
Revisions throughout

\begin{document}
\SweaveOpts{concordance=TRUE}

\title{Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\maketitle

\section{Introduction}

>>>>

\usepackage{Sweave}
\begin{document}
\input{draftfinalreport-concordance}

\title{Draft final report\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\maketitle

\section{Abstract}

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this draft final report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand, and present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on several alternative situations. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one. This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if each measles case in New Zealand continues to infect the same number of secondary cases as we estimate from the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreak dynamics remain the same as those of the average outbreaks since 1997 and future measles cases are few, there is no benefit to additional vaccination campaigns.
\end{itemize}

\section{Background}

As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002. The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved measles elimination.

The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997. Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions. The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{3 July 2014}). In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age. Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

\section{Risk analysis review}

A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups. Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
\begin{itemize}
\item In this section we review the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}.
\end{itemize}

Overall, the review was very thorough. The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}). Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{Section 4.2}) and the sustainability of the national immunisation programme (\emph{Section 5}).

Within the report there are many tables and figures which give considerable detail on the measles situation in New Zealand. Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.

Specific epidemiological details were provided for the 2011--2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6--8}). Similar presentation of the case data are provided for ethnicity (\emph{Figures 9--10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11--13}). Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.

The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ. We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.

We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe. Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving. This is particularly in light of our preliminary modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).

The analyses we believe could further inform the understanding of risk from measles infection are:
\begin{itemize}
\item Multivariate modelling to account for confounding within the univariate analyses.
\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
\item Analyses of changing risk factors through time during outbreaks.
\item Modelling transmission chains in the population to understand effective reproduction number.
\item Development of risk maps to understand measles importation.
\end{itemize}

\subsection{Additional risk analyses}
\label{sub:risk_analyses}

In this section we provide a brief description of the work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles. These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.

\begin{figure}
     \centering
     \includegraphics[width=1\textwidth]{incidence_1997_2014.pdf}
     \caption{Measles incidence from 1997 to 2014}
     \label{fig:incidence1997}
\end{figure}

In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand. To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.

\subsection{Additional risk analyses methods}

For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Note this uses all immigration of foreign nationals, coming for whatever purpose. We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.

%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Total travel numbers (2012)}
%latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
\hline
Australia&$1799655$\tabularnewline
United Kingdom&$ 401737$\tabularnewline
China&$ 322076$\tabularnewline
United States&$ 316058$\tabularnewline
Fiji&$ 151443$\tabularnewline
India&$ 107618$\tabularnewline
Japan&$ 106716$\tabularnewline
Germany&$  96308$\tabularnewline
Korea, Republic of&$  87419$\tabularnewline
France&$  85948$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:immigration12}
\end{table}

%%%%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Lowest national measles vaccine cover (\%, 2012)}
%latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{cover}\tabularnewline
\hline
Equatorial Guinea&$34$\tabularnewline
Somalia&$49$\tabularnewline
Lesotho&$60$\tabularnewline
Central African Republic&$65$\tabularnewline
Papua New Guinea&$67$\tabularnewline
Chad&$69$\tabularnewline
Haiti&$69$\tabularnewline
South Sudan&$70$\tabularnewline
Gabon&$71$\tabularnewline
Yemen&$71$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:cover12}
\end{table}

\begin{table}
\caption{Measles incidence per million (2012)}
%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
\hline
Equatorial Guinea&$1617$\tabularnewline
Nauru&$1100$\tabularnewline
Democratic Republic of the Congo&$1096$\tabularnewline
Somalia&$ 979$\tabularnewline
Djibouti&$ 824$\tabularnewline
Sudan&$ 786$\tabularnewline
Burkina Faso&$ 447$\tabularnewline
Romania&$ 342$\tabularnewline
Ukraine&$ 280$\tabularnewline
Sudan&$ 229$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:incidence12}
\end{table}

\begin{table}
\caption{Risk of measles importation to New Zealand in 2012}
%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
\hline
Australia&$15537152$\tabularnewline
United Kingdom&$13386328$\tabularnewline
Thailand&$ 4774688$\tabularnewline
Malaysia&$ 4495338$\tabularnewline
Indonesia&$ 2216871$\tabularnewline
India&$ 1620172$\tabularnewline
China&$ 1432659$\tabularnewline
Ukraine&$  765096$\tabularnewline
Ireland&$  734128$\tabularnewline
Romania&$  655107$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:risk12}
\end{table}

\begin{table}
\caption{Immigration numbers to New Zealand (2012)}
%latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
\hline
Australia&$809775$\tabularnewline
United Kingdom&$306177$\tabularnewline
China&$256036$\tabularnewline
United States&$194438$\tabularnewline
Japan&$ 86676$\tabularnewline
Germany&$ 83608$\tabularnewline
Korea, Republic of&$ 73459$\tabularnewline
France&$ 71448$\tabularnewline
India&$ 69038$\tabularnewline
Canada&$ 54981$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:immigration12}
\end{table}

%%%%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Lowest national measles vaccine cover (\%, 2012)}
%latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{cover}\tabularnewline
\hline
Equatorial Guinea&$34$\tabularnewline
Somalia&$49$\tabularnewline
Lesotho&$60$\tabularnewline
Central African Republic&$65$\tabularnewline
Papua New Guinea&$67$\tabularnewline
Chad&$69$\tabularnewline
Haiti&$69$\tabularnewline
South Sudan&$70$\tabularnewline
Gabon&$71$\tabularnewline
Yemen&$71$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:cover12}
\end{table}

\begin{table}
\caption{Measles incidence per million (2012)}
%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
\hline
Equatorial Guinea&$1617$\tabularnewline
Nauru&$1100$\tabularnewline
Democratic Republic of the Congo&$1096$\tabularnewline
Somalia&$ 979$\tabularnewline
Djibouti&$ 824$\tabularnewline
Sudan&$ 786$\tabularnewline
Burkina Faso&$ 447$\tabularnewline
Romania&$ 342$\tabularnewline
Ukraine&$ 280$\tabularnewline
Sudan&$ 229$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:incidence12}
\end{table}

\begin{table}
\caption{Risk of measles importation to New Zealand in 2012}
%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
\hline
United Kingdom&$10202161$\tabularnewline
Australia&$ 6991116$\tabularnewline
Malaysia&$ 3226580$\tabularnewline
Thailand&$ 1576414$\tabularnewline
China&$ 1138900$\tabularnewline
India&$ 1039356$\tabularnewline
Indonesia&$  984022$\tabularnewline
Ukraine&$  664315$\tabularnewline
Ireland&$  585413$\tabularnewline
Romania&$  490731$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:risk12}
\end{table}
%%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{New Zealander travel numbers (2012)}
%latex.default(nz, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
\hline
Australia&$989880$\tabularnewline
United States&$121620$\tabularnewline
Fiji&$104720$\tabularnewline
United Kingdom&$ 95560$\tabularnewline
Cook Islands&$ 71960$\tabularnewline
China&$ 66040$\tabularnewline
Samoa&$ 46020$\tabularnewline
Thailand&$ 41100$\tabularnewline
India&$ 38580$\tabularnewline
Canada&$ 20400$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:nztravel12}
\end{table}

%%%%%%%%%%%%%%%%%%%%%

\begin{table}
\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
%latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lr}
\hline\hline
\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
\hline
Australia&$8546036$\tabularnewline
Thailand&$3198274$\tabularnewline
United Kingdom&$3184166$\tabularnewline
Malaysia&$1268758$\tabularnewline
Indonesia&$1232849$\tabularnewline
India&$ 580816$\tabularnewline
China&$ 293759$\tabularnewline
Samoa&$ 243492$\tabularnewline
Philippines&$ 241740$\tabularnewline
Nepal&$ 188450$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:nzrisk12}
\end{table}

%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%
\subsection{Risk analyses methods}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago. We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of risk of measles infection.

The University of Otago denominator data provided were not to the same detail as the ESR case data. Notably, the denominator age data were categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories. The ethnicity denominator data were not Prioritized Ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes. After discussions with the University of Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.

With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories. Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).

\begin{figure}[h!]
\begin{center}
\includegraphics{draftfinalreport-015}
\end{center}
\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
\label{fig:ageinyears}
\end{figure}

This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 and NZDep 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories. For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}).

\begin{table}\tiny
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
\hline
1-5&0-2&Asian&$ 11$\tabularnewline
6-10&0-2&Asian&$  8$\tabularnewline
1-5&3-5&Asian&$  1$\tabularnewline
1-5&6-17&Asian&$ 11$\tabularnewline
6-10&6-17&Asian&$  5$\tabularnewline
1-5&18-24&Asian&$  3$\tabularnewline
6-10&18-24&Asian&$  5$\tabularnewline
1-5&25+&Asian&$ 10$\tabularnewline
6-10&25+&Asian&$ 13$\tabularnewline
1-5&0-2&European&$ 83$\tabularnewline
6-10&0-2&European&$ 64$\tabularnewline
1-5&3-5&European&$ 42$\tabularnewline
6-10&3-5&European&$ 17$\tabularnewline
1-5&6-17&European&$219$\tabularnewline
6-10&6-17&European&$ 80$\tabularnewline
1-5&18-24&European&$ 34$\tabularnewline
6-10&18-24&European&$ 36$\tabularnewline
1-5&25+&European&$ 78$\tabularnewline
6-10&25+&European&$ 51$\tabularnewline
1-5&0-2&Maori&$ 18$\tabularnewline
6-10&0-2&Maori&$ 48$\tabularnewline
1-5&3-5&Maori&$  7$\tabularnewline
6-10&3-5&Maori&$ 11$\tabularnewline
1-5&6-17&Maori&$ 19$\tabularnewline
6-10&6-17&Maori&$ 92$\tabularnewline
1-5&18-24&Maori&$  5$\tabularnewline
6-10&18-24&Maori&$  8$\tabularnewline
1-5&25+&Maori&$  2$\tabularnewline
6-10&25+&Maori&$  6$\tabularnewline
6-10&0-2&MLA&$  3$\tabularnewline
6-10&3-5&MLA&$  1$\tabularnewline
6-10&6-17&MLA&$  1$\tabularnewline
6-10&18-24&MLA&$  2$\tabularnewline
6-10&25+&MLA&$  6$\tabularnewline
1-5&0-2&Pacific&$  5$\tabularnewline
6-10&0-2&Pacific&$ 58$\tabularnewline
6-10&3-5&Pacific&$  3$\tabularnewline
1-5&6-17&Pacific&$  5$\tabularnewline
6-10&6-17&Pacific&$ 22$\tabularnewline
1-5&18-24&Pacific&$  1$\tabularnewline
6-10&18-24&Pacific&$  8$\tabularnewline
1-5&25+&Pacific&$  2$\tabularnewline
6-10&25+&Pacific&$ 11$\tabularnewline
1-5&0-2&None&$  3$\tabularnewline
1-5&3-5&None&$  1$\tabularnewline
1-5&6-17&None&$  3$\tabularnewline
6-10&6-17&None&$  4$\tabularnewline
1-5&18-24&None&$  2$\tabularnewline
6-10&18-24&None&$  1$\tabularnewline
1-5&25+&None&$  5$\tabularnewline
6-10&25+&None&$  3$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:case}
\end{table}

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table~\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}[h!]
\begin{center}
\includegraphics{draftfinalreport-023}
\end{center}
\caption{Per capita rates of measles infections broken down by ethnicity, see Table~\ref{table:percap} for details}
\label{fig:percap}
\end{figure}

\begin{table}\tiny
\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
\hline
1-5&0-2&Asian&$   6094$&$ 11$&$0.0018$\tabularnewline
6-10&0-2&Asian&$   6806$&$  8$&$0.0012$\tabularnewline
1-5&3-5&Asian&$   6094$&$  1$&$0.0002$\tabularnewline
6-10&3-5&Asian&$   6806$&$  0$&$0.0000$\tabularnewline
1-5&6-17&Asian&$  33918$&$ 11$&$0.0003$\tabularnewline
6-10&6-17&Asian&$  28905$&$  5$&$0.0002$\tabularnewline
1-5&18-24&Asian&$  22917$&$  3$&$0.0001$\tabularnewline
6-10&18-24&Asian&$  34107$&$  5$&$0.0001$\tabularnewline
1-5&25+&Asian&$  96357$&$ 10$&$0.0001$\tabularnewline
6-10&25+&Asian&$  98715$&$ 13$&$0.0001$\tabularnewline
1-5&0-2&European&$  57872$&$ 83$&$0.0014$\tabularnewline
6-10&0-2&European&$  45445$&$ 64$&$0.0014$\tabularnewline
1-5&3-5&European&$  57872$&$ 42$&$0.0007$\tabularnewline
6-10&3-5&European&$  45445$&$ 17$&$0.0004$\tabularnewline
1-5&6-17&European&$ 264330$&$219$&$0.0008$\tabularnewline
6-10&6-17&European&$ 182937$&$ 80$&$0.0004$\tabularnewline
1-5&18-24&European&$ 107649$&$ 34$&$0.0003$\tabularnewline
6-10&18-24&European&$ 117840$&$ 36$&$0.0003$\tabularnewline
1-5&25+&European&$1001916$&$ 78$&$0.0001$\tabularnewline
6-10&25+&European&$ 724317$&$ 51$&$0.0001$\tabularnewline
1-5&0-2&Maori&$  10003$&$ 18$&$0.0018$\tabularnewline
6-10&0-2&Maori&$  30104$&$ 48$&$0.0016$\tabularnewline
1-5&3-5&Maori&$  10003$&$  7$&$0.0007$\tabularnewline
6-10&3-5&Maori&$  30104$&$ 11$&$0.0004$\tabularnewline
1-5&6-17&Maori&$  40461$&$ 19$&$0.0005$\tabularnewline
6-10&6-17&Maori&$ 116640$&$ 92$&$0.0008$\tabularnewline
1-5&18-24&Maori&$  15360$&$  5$&$0.0003$\tabularnewline
6-10&18-24&Maori&$  48495$&$  8$&$0.0002$\tabularnewline
1-5&25+&Maori&$  71217$&$  2$&$0.0000$\tabularnewline
6-10&25+&Maori&$ 192729$&$  6$&$0.0000$\tabularnewline
1-5&0-2&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&0-2&MLA&$   1290$&$  3$&$0.0023$\tabularnewline
1-5&3-5&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&3-5&MLA&$   1290$&$  1$&$0.0008$\tabularnewline
1-5&6-17&MLA&$   2991$&$  0$&$0.0000$\tabularnewline
6-10&6-17&MLA&$   4539$&$  1$&$0.0002$\tabularnewline
1-5&18-24&MLA&$   1710$&$  0$&$0.0000$\tabularnewline
6-10&18-24&MLA&$   3078$&$  2$&$0.0006$\tabularnewline
1-5&25+&MLA&$   8028$&$  0$&$0.0000$\tabularnewline
6-10&25+&MLA&$  10335$&$  6$&$0.0006$\tabularnewline
1-5&0-2&Pacific&$   2093$&$  5$&$0.0024$\tabularnewline
6-10&0-2&Pacific&$  13124$&$ 58$&$0.0044$\tabularnewline
1-5&3-5&Pacific&$   2093$&$  0$&$0.0000$\tabularnewline
6-10&3-5&Pacific&$  13124$&$  3$&$0.0002$\tabularnewline
1-5&6-17&Pacific&$   8541$&$  5$&$0.0006$\tabularnewline
6-10&6-17&Pacific&$  51183$&$ 22$&$0.0004$\tabularnewline
1-5&18-24&Pacific&$   3972$&$  1$&$0.0003$\tabularnewline
6-10&18-24&Pacific&$  22098$&$  8$&$0.0004$\tabularnewline
1-5&25+&Pacific&$  18492$&$  2$&$0.0001$\tabularnewline
6-10&25+&Pacific&$  91533$&$ 11$&$0.0001$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:percap}
\end{table}

For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and then simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

\begin{equation} \label{eq:reg}
 log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a * x_e)+ log(population)  + \epsilon
  \end{equation}
Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.

\subsection{Additional risk analyses results}

Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}. However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}). Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand. These data are shown in Table \ref{table:risk12}. The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{np1.pdf}
     \caption{Non-New Zealander travel}
     \label{fig:immigration12}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{nznp1.pdf}
     \caption{New Zealand travellers 2012}
     \label{fig:nztravellers}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{totnp1.pdf}
     \caption{Total travellers 2012}
     \label{fig:tottravellers}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{draftfinalreport-028}
\end{center}
\caption{Measles incidence per million 2012}
\label{fig:incidence12}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{draftfinalreport-029}
\end{center}
\caption{Measles vaccination cover (\%) 2012 }
\label{fig:cover12}
\end{figure}

\begin{figure}[h!]
 \centering
 \includegraphics[width=0.8\textwidth]{np4.pdf}
\caption{Risk map from measles incidence and immigration rates}
\label{fig:risk12}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{nznp4.pdf}
     \caption{Risk map from measles incidence and New Zealander travel rates}
     \label{fig:nzrisk12}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.8\textwidth]{totnp4.pdf}
     \caption{Risk map from measles incidence and total travel rates}
     \label{fig:totrisk12}
\end{figure}

%%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
\begin{figure}[h!]
\begin{center}
\includegraphics{draftfinalreport-030}
\end{center}
\caption{Trend in global per capita measles incidence}
\label{fig:trendincidence}
\end{figure}

\begin{figure}
     \begin{center}
     \includegraphics[width=1\textwidth]{nzers.pdf}
     \end{center}
     \caption{Trends in travel}
     \label{fig:travel}
\end{figure}

Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}). This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.  Of note, however, is the clear seasonality in immigration (Figure~\ref{fig:trendimmigration}).  This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
